### 2016-04-22 manuscript-body.tex
Get manuscript to at least compile

\subsection{Introduction}\label{introduction-2}

Although imaginative and carefully designed, the Harvard course taught
that there is only one right way to approach the computer, a way that
emphasizes control through structure and planning. There are many
virtues to this computational approach (it certainly makes sense when
dividing the labor on a large programming project), but Lisa and Robin
have intellectual styles at war with it. Lisa says she has ``turned
herself into a different kind of person'' in order to perform, and Robin
says she has learned to ``fake it.'' Although both women are able to get
good grades in their programming course, they represent casualties of
this war. Both deny who they are in order to succeed. (Turkle \& Papert,
1991, p. 165)

Over 20 years ago Sherry Turkle and Seymour Papert called for
epistemological pluralism in computing education. Their argument --- in
my view ---had multiple parts:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Programming a computer involved a shift in how one thought about the
  nature knowledge from a propositional \emph{what is} to an imperative
  \emph{how to} (Abelson \& Sussman, 1996; Papert, 1980).
\item
  Historically, there was a tendency to treat computers and the act of
  programming them as an extension of the same Western logico-deductive
  knowledge tradition that fueled science.
\item
  Recent scholarship, often at the intersection of post-structuralism
  and social studies of science, had been troubling the idea that
  knowledge-work in science proceeded solely by way of logical
  deductions (Keller, 1983; Latour, 1987; Traweek, 1988)
\item
  Data from professionals suggested programming, too, didn't proceed
  solely by way of the structured-planning approach emergent from
  Western logico-deductive knowledge traditions. There were many ways of
  knowing and constructing knowledge in programming.
\item
  A viewpoint of \emph{epistemological pluralism} --- embracing multiple
  kinds and ways of knowing --- was informing the field of science
  studies but not, to a large degree, how education researchers thought
  about or taught computing.
\item
  The lack of support for diverse ways of knowing in computing
  classrooms demonstrably hurt students.
\end{enumerate}

To Turkle and Papert (1991), the result of this culturally-rooted
epistemological tension was a war. More specifically, it was a war where
students of computing were the casualties, and the aggressor was a
distributed failure to recognize, embrace, and support diverse ways of
knowing. While I think their strong-form characterization of a ``war''
no longer applies, their consideration of epistemological issues is
still relevant two decades later. An understanding that computing does
in fact involve diverse ways of knowing should continue to inform
research and discussions on learning

This paper tries to model what directs and sustains students'
in-the-moment activity when they program. Its focus is on early-stage
program design; analyzing how what students say, do, write, and gesture
\emph{even before they type code} can help us improve theories of
cognition and activity in learning programming. Using a ``revelatory
case study'' (Yin, 2009) of two students in an introductory programming
course, I argue the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Students' early-stage design activity reveals patterns outside the
  explanatory scope of misconception-based accounts of cognition.
\item
  For a subset of phenomena, we can recast students' productive
  capacities and their difficulties in terms of epistemological stances.
\item
  Such recastings are analytically powerful. Beyond Rebecca and Lionel,
  they could potentially explain the diversity of practices I saw other
  students take up.
\item
  As evidenced by work in science and math education, dynamic
  epistemological models can offer a lens for reforming assessment and
  instruction.
\end{enumerate}

\subsection{Literature Review}\label{literature-review}

I begin by exploring the historical legacy of Turkle and Papert's (1991)
remarks on epistemological pluralism in computing. I'll then describe
what I see as a well-intentioned obstacle: the research-based
preoccupation with students' misconceptions in computing education.
These two stra.

\subsubsection{Computing culture should support a diversity of ways of
knowing}\label{computing-culture-should-support-a-diversity-of-ways-of-knowing}

Turkle \& Papert (1991) describe a tension between individuals and a
kind of cultural collective. In this tension, individual students and
their personally-identified ways of knowing were in conflict with the
larger culture of computing and its manifested ways of knowing. Lisa,
for instance, was a Harvard student who identified as a poet:

Lisa experiences language as transparent, she knows where all the
elements are at every point in the development of her ideas. She wants
her relationship to computer language to be similarly transparent. When
she builds large programs she prefers to write her own smaller
``building block'' procedures even though she could use prepackaged ones
from a program library; she resents the latter's opacity. (Turkle \&
Papert, 1991, p. 164)

Turkle and Papert connect Lisa's programming practices to Lisa's
personal view of knowledge. She writes building block procedures because
of a continuity between herself as knower-of-her-own-poetry and that of
knower-of-her-own-programs. Across both contexts, Lisa \emph{resents
opacity} because it is an obstacle to her knowing. But, that resentment
slows her progress on projects in the class.

In part because of her commitments to transparency and the difficulty
they created, Lisa changed her approach to programming. She ultimately
had to, in her words, ``be a different kind of person with the machine''
(Turkle \& Papert, 1991, p. 164). Turkle and Papert explain:

She had been told that the ``right way'' to do things was to control a
program through planning and black­boxing, the technique that lets you
exploit opacity to plan something large without knowing in advance how
the details will be managed. Lisa recognized the value of these
techniques --- for someone else. She struggled against using them as the
starting points for her learning. Lisa ended up abandoning the fight,
doing things ``their way,'' and accepting the inevitable alienation from
her work. (Turkle \& Papert, 1991, p. 164)

The phrase ``inevitable alienation'' is particularly damning. One might
accuse Turkle and Papert (1991) of hyperbole, but research from other
disciplines suggests they're not overstating the case. Identity
alienation for epistemological reasons --- or some variation of them ---
is playing out in other subjects as well.

Connections and tensions between identities, learning, and personal ways
of knowing are not unique to computing education. Wortham (2006), for
example, detailed instances of social identification and academic
identity as \emph{jointly emerging} in a combined high school English
and History class. Nasir and colleagues explored how different learning
environments afford access to identity, domain practices, and
self-expression (Nasir \& Cooks, 2009; Nasir \& Hand, 2008). To single
out one study, Nasir \& Hand (2008) followed players ``from the
{[}basketball{]} court to the classroom.'' They found the court afforded
players a sense of role on the team, chances to be expressive through
play, and ``access to the domain {[}of basketball{]} as a whole'' (Nasir
\& Hand, 2006, p. 147). Comparatively, the mathematics classroom offered
little (if any) sense that students were on a team, a much narrower role
for them as having either right or wrong answers, and no strong sense of
self-expression or creativity through activity. These markedly different
contexts led not only to differences in kinds of activity but, Nasir and
Hand (2006) argue, divergent \emph{practice-linked identities} for the
players. In a sense, players had a different identity on the court than
they did in the mathematics classroom because who they were was strongly
coupled to a sense of their relationship with a practice.

Boaler's work on learning and identity in mathematics classrooms (Boaler
\& Greeno, 2000; Boaler, 1998, 2000, 2002) offers resonant findings. In
particular, Boaler (2000) strongly echoes the student voices in Turkle
and Papert (1991). Where Turkle and Papert's aggressor is the dominant
computing culture, Boaler describes a beast with similar effects in the
form of ``school mathematics'' --- the \emph{monotonous},
\emph{meaningless} (Boaler, 2002, pp. 383--384) imposition of routinized
mathematical procedures in ways that seem far removed from the real
world. Of school mathematics, Boaler writes:

School mathematics, for many of them, was \emph{of another world} and to
fully engage in that world, students needed to suspend their knowledge
of the real world, suppress their desire to interact with others, and
strive to reproduce standard procedures that held little meaning for
them. (Boaler, 2002, p. 392 emphasis in original)

Schoenfeld (1988, 1991) had already argued school mathematics could and
did depart substantially from the core of mathematical thinking. Boaler
established that such school mathematics could alienate learners
(Boaler, 1998, 2000) and influence their feelings about pursuing
mathematics after secondary school (Boaler \& Greeno, 2000).

Most recently, Danielak, Gupta, and Elby (in press) extended those
findings in engineering education. In their 3 ½ year case study of an
electrical engineering undergraduate named ``Michael,'' those authors
argued Michael's practice of sense-making in engineering coursework was
strongly coupled to his identity. Michael's passion for deeply
understanding concepts, much like Lisa's (Turkle \& Papert, 1991) desire
to understand every element of her program, did not mesh neatly with the
larger culture of his courses. But, because Michael's practice of
sense-making entwined with his identity, forces that pushed against
sense-making alienated him. That tension was particularly pointed when
Michael described his father's resistance to Michael's way of thinking:

{[}My dad said{]} ``you're just an undergraduate. Nobody expects
undergraduates to understand how anything works. That's why you go to
graduate school.'' I was like ``look, you know. I'm gonna be an unhappy
person if I have to....I have life goals other than to just get a good
grade on the exam. Other things are important to me.'' (Danielak et al.,
in press)

And, much like Turkle and Papert's (1991) Lisa, Michael learned to
curtail parts of his identity to get by in class:

I think the reason {[}pursuing deep learning{]} actually hasn't affected
my GPA is because I view learning as a hobby. So, as with any hobby, you
shouldn't let it interfere your GPA. But it \emph{is} one of my hobbies,
and I \emph{do} enjoy learning, I just---up to the point where I get my
grades done \{raises eyebrow\}. (Danielak et al., in press)

Ultimately, research suggests the identity-epistemology tensions Turkle
and Papert (1991) described in computing resonates with those found in
related disciplines. And, what such accounts --- particularly those of
Boaler and others (Boaler \& Greeno, 2000; Boaler, 1998, 2000, 2002;
Danielak et al., in press) --- have in common is a culturally-sanctioned
``right way'' (Turkle \& Papert, 1991, p. 164) of thinking or knowing
that alienates some students. Avowedly such ways of knowing have use and
value. Turkle \& Papert themselves acknowledge ``there are many
virtues'' to black-boxing and top-down design in programming (Turkle \&
Papert, 1991, p. 165). But,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  When a preponderance of results from science studies suggests a
  plurality of ways of knowing among practicing scientists and
  programmers, and
\item
  When research from education shows that the inflexible imposition of
  one way of knowing above all else is alienating students, then
\item
  It's worth reconsidering whether a narrow sense of what counts as
  knowing might still be hindering improvements in computing education.
\end{enumerate}

In the next section, I turn to misconceptions research in computing
education. In keeping with observations 1--3,

@@

In the past three decades, educational research has had a marked focus
on students' misconceptions in programming. It's a focus with a sensible
origin. Students get things wrong in programming --- often
systematically so --- and in ways that seem resistant to instruction.
The cause of those errors is theorized to be something cognitive,
whether it's a ``bug'' (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
rule'' (Fleury, 1991, 2000). Instruction should try to identify,
address, and correct these misconceptions (Clancy, 2004) because they
can represent barriers to learning.

@@

\begin{itemize}
\item
  Objects in object-oriented programming (Holland, Griffiths, \&
  Woodman, 1997)
\item
  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
  Paul \& Vahrenhold, 2013)
\item
  Programming statements in BASIC (Bayman \& Mayer, 1983)
\item
  Programming in Java (Fleury, 2000)
\item
  Programming in Pascal (Fleury, 1993)
\item
  Parameter-passing (Fleury, 1991)
\item
  Arrays in Java (Kaczmarczyk et al., 2010)
\item
  Objects in Java (Kaczmarczyk et al., 2010)
\item
  Algorithms and computational complexity (Trakhtenbrot, 2013)
\item
  Boolean logic (Herman et al., 2008)
\item
  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
\item
  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
\item
  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
\item
  The correctness of programs (Kolikant \& Mussai, 2008)
\end{itemize}

@@

\subsection{Acknowledgments}\label{acknowledgments}

Blinded for review.

\clearpage

\bibliographystyle{acm}
\bibliography{bibliography}

>>>>

@@

In the past three decades, educational research has had a marked focus
on students' misconceptions in programming. It's a focus with a sensible
origin. Students get things wrong in programming --- often
systematically so --- and in ways that seem resistant to instruction.
The cause of those errors is theorized to be something cognitive,
whether it's a ``bug'' \cite{pea_buggy_1987} (Pea et al., 1987; Pea, 1986; VanLehn, 1990) a
``misconception'' (Bayman \& Mayer, 1983; Bonar \& Soloway, 1985;
Clancy, 2004; Gal-Ezer \& Zur, 2004; Herman et al., 2008; Kaczmarczyk et
al., 2010), a ``belief'' (Fleury, 1993) or a ``student-constructed
rule'' (Fleury, 1991, 2000). Instruction should try to identify,
address, and correct these misconceptions (Clancy, 2004) because they
can represent barriers to learning.

@@

\begin{itemize}
\tightlist
\item
  Objects in object-oriented programming (Holland, Griffiths, \&
  Woodman, 1997)
\item
  Algorithms and data structures (Danielsiek, Paul, \& Vahrenhold, 2012;
  Paul \& Vahrenhold, 2013)
\item
  Programming statements in BASIC (Bayman \& Mayer, 1983)
\item
  Programming in Java (Fleury, 2000)
\item
  Programming in Pascal (Fleury, 1993)
\item
  Parameter-passing (Fleury, 1991)
\item
  Arrays in Java (Kaczmarczyk et al., 2010)
\item
  Objects in Java (Kaczmarczyk et al., 2010)
\item
  Algorithms and computational complexity (Trakhtenbrot, 2013)
\item
  Boolean logic (Herman et al., 2008)
\item
  The efficiency of algorithms (Gal-Ezer \& Zur, 2004)
\item
  The Build-Heap algorithm (Seppälä, Malmi, \& Korhonen, 2006)
\item
  Hashtables (Patitsas, Craig, \& Easterbrook, 2013)
\item
  The correctness of programs (Kolikant \& Mussai, 2008)
\end{itemize}

@@
